---
title: 'Introduction to R'
author: 'Caroline Hutchings'
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document: 
    code_folding: show
    theme: 'flatly'
    toc: true
    toc_float: true

---
## What is R? {.tabset}
**AI Definition:**

R is a programming language and free software environment primarily used for statistical computing and graphics. It is widely utilized among statisticians, data scientists, and researchers for developing statistical software and data analysis.

### Key Features of R:
- **Built-in Statistics Functions**
  * Provides a variety of statistical tests, such as t-tests, ANOVA, etc.
- **Visualization Capabilities**
  * Tools like `ggplot2` facilitate high-quality visualizations.
- **Expanded Capabilities through Added Libraries**
  * Thousands of packages are available on CRAN.
- **Integration with Other Tools, such as SQL**
  * Connects with databases for efficient data manipulation.
  
### Applications of R:
- **Data Science**
- **Statistics**
- **Academic Research**
- **Business Analytics**
- **Bioinformatics**
- **Finance**

## RStudio Environment

- **Source Pane**
  - Top left corner
  - Used for writing and editing R scripts and RMarkdown documents.
  
- **Console Pane**
  - Bottom left corner
  - Executes R commands interactively.

- **Environment/History Pane**
  - Top right corner
  - **Environment Tab**: Displays variables and objects in memory.
  - **History Tab**: Shows command history for the current session.

- **Files/Plots/Packages/Help/Viewer Pane**
  - Bottom right corner
  - **Files Tab**: Navigate and manage files in the working directory.
  - **Plots Tab**: Displays visual outputs and allows interaction with plots.
  - **Packages Tab**: Lists installed packages; enables installation and loading.
  - **Help Tab**: Access to R documentation and function help.
  
## What is RMarkdown?

- RMarkdown is an authoring format that allows you to combine text, code, and output in a single document.

## Key Features of RMarkdown
- **Output Control**: The use of chunk options like message=FALSE, echo=FALSE, warning=FALSE, and results=FALSE enables users to suppress messages, code display, warnings, and results in the final document, allowing for a cleaner and more focused presentation of content.
- **Multiple Output Formats**: Create documents in formats such as HTML, PDF, Word, and PowerPoint.
- **Easy Text Formatting**: Text can be formatted using Markdown Syntax.
- **Integration with R Packages**: Supports R packages like **'ggplot2** for visualization, **'knitr'** for report generation and **'dplyr'** for data manipulation.
- **Interactive Features**: Interactive elements, such as plots users can zoom, can be included in reports.

## How to Use RMarkdown

- **Creating an RMarkdown Document**:
  - In RStudio, navigate to **File -> New File -> R Markdown…** to create a new document.

- **Writing and Editing**:
  - Write your text using Markdown syntax.
  - Insert R code chunks by navigating to **+C** and add an R chunk.
  - Structure your document with headers for organization.

- **Knit the Document**:
  - Click the **"Knit"** button to generate the output in your desired format (HTML, PDF, Word, PowerPoint, etc.).

## Format Header in RMarkdown

#### Key Components of the Header

- **title**: Sets the title of your document.
- **author**: Specifies the author's name.
- **date**: Format the date automatically using R code.
- **output**: Defines the output format (e.g., PDF, HTML, PowerPoint).

## Example Header

```yaml
---
title: "Your Presentation Title"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    theme: "cosmo"  
---

```
## Vectors

- Create a vector of numbers using assignment operator
```{r message=FALSE, echo=FALSE, warning=FALSE, results=FALSE}
nums <- c(1,2,3) #assignment operator
nums
```
- Create a vector of strings
```{r}
week <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
```
- Create a vector with a range of numbers
```{r}
seq(1,10)
```
- Length of vector
```{r}
length(week)
```
## Working with Vectors

- Create a named vector
```{r}
temps <- c('Mon'=88, 'Tue'=93, 'Wed'=95, 'Thu'=91, 'Fri'=86)
temps
```
- Retrieve names of vector
```{r}
names(temps)
```
- Add values to a vector
```{r}
temps <- append(temps, c('Sa'=81, 'Su'=78))
temps
```
## Indexing

- Identify item at specific index
```{r}
week[1]
```
- All items except at specified index
```{r}
week[-2]
```
- All items in vector of indexes
```{r}
week[c(1,3,5)]
```


## Working with Vector Operators

- Relational Operators
```{r}
temps == 81
temps > 88
temps != 81
```

- Write code to shorten the names of temps 
```{r }
names(temps) <- c('M', 'T', 'W', 'Th', 'F')
temps
```

- Write code to select temperature of Friday
```{r}
temps['F']
temps[5]
```
## Dataframes

- Create a dataframe from vectors
```{r}
months <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
num_days <- c(31,28,31,30,31,30,31,31,30,31,30,31)
seasons <- c('W', 'W', 'Sp', 'Sp', 'Sp', 'Su', 'Su', 'Su', 'F', 'F', 'F', 'W')
calendar_df <- data.frame(Month = months, NumDays = num_days, Season=seasons)
calendar_df

```
## Dataframe Manipulations

- Add rows to a dataframe
```{r}
calendar_df <- rbind(calendar_df, list('NewJan', 31, 'Bonus'), list('NewFeb', 29, 'Bonus'))
calendar_df
```
- Add columns to a dataframe
```{r}
avg_temp <- c(52.3, 56.1, 62, 68.9, 75.9, 80.7, 81.9, 82.7, 78.5, 70.7, 61.1, 54.3, 0, 0)
calendar_df <- cbind(calendar_df, AvgTemp = avg_temp)
calendar_df
```
## Logical Operators

- Find Months with 30 days that are warmer than 70 degrees.
- Find Months that have a temperature lesss than 70 degrees or are summer months.
```{r}
calendar_df$AvgTemp >=70 & calendar_df$NumDays==30 #AND
calendar_df$AvgTemp <=70 | calendar_df$Season=='Su'
```
## Dataframe Application

- Find proportion of months with 31 days that are above 70 degrees.
- Find proportion of months that have less than 31 days or are above 70 degrees.

## Tidyverse

- Tidyverse is a set of packages that facilitate data science tasks in R. The main focus is on providing tools for data manipulation, exploration, and visualization in a cohesive and consistent way.

- **Tidy Data**: Where each variable is a column, each observation is a row, and each type of observational unit forms a table. This structure makes it easier to work with and visualize data.
  - **Tibble**: modern, user-friendly version of data frames in R that adheres to the principles of tidy data. 
- **Modularity** : Allowing users to load only the packages they need for specific tasks, making the workflow more efficient.

## Core Components of Tidyverse

- **ggplot2**: 
  - Purpose: Data visualization
  - Features: Provides a coherent system for creating graphs
  - Usage: Create complex visualizations by layering
- **dplyr**:
  - Purpose: Data manipulation
  - Functions: select, filter, etc. to manipulate tibbles.
- **tidyr**: 
  - Purpose: Data tidying
  - Functions: pivot_longer, pivot_wider, etc. help reshape data into a tidy format.
- **readr**:
  - Purpose: Data import
  - Functions: read_csv, write_csv, read_delim, etc. imports data into a tibble by default.
- **readxl**:
  - Purpose: Import excel data
  - Functions: read_excel imports data into a tibble by default.
  
## Essential Tidyverse Functions

### Key Functions in dplyr

- **mutate()**
  - Adds or modifies columns in a data frame.

- **filter()**
  - Subsets rows based on specified conditions.

- **select()**
  - Chooses specific columns from a data frame.

- **summarize()**
  - Generates summary statistics for variables.

- **arrange()**
  - Sorts data by one or more columns.

- **group_by()**
  - Groups data for analysis, facilitating operations like summarization.
  
- **piping (%>%)**
  - Allows for chaining multiple dplyr functions together.  Passes the output of one function directly to the next, enhancing readability and efficiency in data manipulation.
  
## Dataframe Vs Tibble

Tibbles are designed for a more user-friendly experience in managaing and analyzing data, especially larger data sets.  To check if tibble -> is_tibble(df)

#### Tibble
- Printing: Displays first 10 rows
- Column Types: Displays column types alongside data
- Subsetting: Maintains tibble structure when subsetting
- Design: Use with Tidyverse
- Handling of Data: Enforces stricter rules

#### Dataframe
- Printing: Displays all rows
- Column Types: Does not display column types
- Subsetting: Converts to vector when single row selected
- Design: Use with base R
- Handling of Data: More flexible

```{r}
library(tidyverse)
#convert a dataframe to a tibble
calendar_tibble <- as_tibble(calendar_df)
is_tibble(calendar_tibble)
is_tibble(calendar_df)
```


## Load Data Set

- Load data using Kaggle API Token

```{r}
#install.packages("httr")
#install.packages("jsonlite")
library(httr)
setwd("C:/Users/cbhut/.kaggle/")
list.files()

system("kaggle datasets download -d zeeshier/student-admission-records")
unzip("student-admission-records.zip") 
student_admission <- read_csv("student_admission_record_dirty.csv")  
is_tibble(student_admission)

```
- Download data to computer and load data set
```{r}
setwd("C:/Users/cbhut/Downloads/") #set to location of file
student_admission <- read_csv("student_admission_record_dirty.csv") 
is_tibble(student_admission)
student_admission
```
## View Data in Spreadsheet Format

- View data in interactive window
```{r}
View(student_admission)
```
- Features of the View(df) :
  - Column Headings : Each column in the data frame is represented as a header in the viewer.
  - Data Rows : Each row of data corresponds to an observation and is displayed in a horizontal format.
  - Scrolling : You can scroll through both the rows and columns if your data frame is larger than the viewing area.
  - Sorting : You can click on column headers to sort the data in ascending or descending order.
  - Filtering : Depending on the interface, you might be able to filter data directly through the viewer.
  
## Missing Data

- Check for missing data
```{r}
#count of rows that have missing name
sum(is.na(student_admission$Name))

#count of NA values for all columns
student_admission %>% 
  apply(MARGIN = 2, function(col) sum(is.na(col)))

#count of rows with one or more missing values
sum(!complete.cases(student_admission))

#drop rows with one or more missing values
student_admission_dropna <- student_admission %>% 
  filter(complete.cases(student_admission))
nrow(student_admission_dropna)
```
- Replace missing values with 'Unknown'
```{r}
#replace Name with 'Unknown'
student_admission_replace_na <- student_admission %>% 
  mutate(Name = replace_na(Name, 'Unknown'))

nrow(student_admission_replace_na)

sum(!complete.cases(student_admission_replace_na))


#replace Name and Admission Status with 'Unknown'
student_admission_replace_na_two <- student_admission %>% 
  mutate(across(c(Name, `Admission Status`),~ replace_na(.,'Unknown')))

nrow(student_admission_replace_na_two)

sum(!complete.cases(student_admission_replace_na_two))

```
## Basketball Example 

An example from Murach's R for Data Analysis Book. Install packages and libraries
```{r}
#install.packages("RJSONIO")
library(tidyverse)
library(RJSONIO)


```
## Load Data from JSON File
```{r}
# Download the JSON data
url <- "https://www.murach.com/python_analysis/shots.json"
download.file(url, "shots.json")

# Load the JSON data into R
json_data <- fromJSON("shots.json")
columnname <- json_data[["resultSets"]][[1]][["headers"]]
rows_data <- json_data[["resultSets"]][[1]][["rowSet"]]

#Create empty dataframe
shots <- data.frame()

#Loop through each row and add to dataframe
for (row in rows_data){
  shots <- rbind(shots, row)
}

#Set column names
names(shots) <- columnname

#Change Dataframe to tibble
shots_tibble <- as_tibble(shots)
View(shots_tibble)

```

## Manipulate Data

### Examine Count of Unique Values
```{r}
#lapply is list apply and apply is column/matrix apply
apply(X=shots_tibble, MARGIN = 2, FUN = unique) %>% 
  lapply(FUN = length) %>% str()
```
### Examine Unique Values of Shot Flag
```{r}
unique(shots_tibble$SHOT_MADE_FLAG)
```
### Select and Rename Columns
```{r}
shots_tibble <- shots_tibble %>% 
  select(GAME_DATE, SHOT_TYPE, EVENT_TYPE, SHOT_MADE_FLAG, LOC_X, LOC_Y, SHOT_ZONE_BASIC)

names(shots_tibble) <- c("GameDate", "ShotType", "EventType", "ShotMadeFlag", "LocX", "LocY", "Zone")
```
### Change Data Types 
```{r}
shots_tibble <- shots_tibble %>% 
  mutate(GameDate = as.Date(GameDate, format = "%Y%m%d"),
         ShotMadeFlag = as.logical(ShotMadeFlag))

shots_tibble
```
### Create Season Variable and Add to Tibble
```{r}
#Create Season Variable
get_season <- function(row){
  month <- as.integer(format(as.Date(row["GameDate"]), "%m"))
  year <- as.integer(format(as.Date(row["GameDate"]), "%Y"))
  season <- ""
  if(month > 6){
    season <- str_c(year, "-", year+1)
  }
    else {
      season <- str_c(year-1, "-", year)
    }
  return(season)
                      
}

#Add to tibble
shots_tibble <- shots_tibble %>% 
  mutate(Season = apply(X=shots_tibble, MARGIN = 1, FUN = get_season))
```
### Create Points Columns
```{r}
#Create variable points
shots_tibble <- shots_tibble %>% 
  mutate(Points = ifelse(ShotMadeFlag ==TRUE, 
                         ifelse(ShotType == "3PT Field Goal", 3, 2),
                         0))

#Create three summary columns
shots_tibble <- shots_tibble %>% 
  group_by(GameDate) %>% mutate(PointsPerGame = sum(Points),
                                AttemptedPerGame = n(),
                                MadePerGame = sum(ShotMadeFlag))

View(shots_tibble)
```
### Shot Statistics per Season
```{r}
shot_season <- shots_tibble %>% 
  group_by(Season) %>% 
  summarize(MeanMade = mean(MadePerGame),
            MeanAttempted = mean(AttemptedPerGame),
            MeanPoints = mean(PointsPerGame))

#Shooting percentage per season
shots_pct_season <- shots_tibble %>% 
  group_by(Season) %>% 
  summarize(Made = sum(ShotMadeFlag),
            Attempted = n()) %>% 
  mutate(SeasonPct = (Made / Attempted),
         CumulativePct = (cumsum(Made) / cumsum(Attempted)))

shots_pct_season
```
## On Your Own - Airport Example

